Fast and optimal centroiding of faint stars

### 2014-12-12 ms/figure.tex
Update figure.tex

If we apply this method to every column and row of the 3$\times$3 patch around the brightest pixel of the smoothed image, we will find three local maxima for three rows $\{(x_{-},-1)$, $(x_{0},0)$, $(x_{+},1)\}$, and three local maxima for three columns $\{(-1,y_{-})$, $(0,y_{0})$, $(1,y_{+})\}$. It is important to note that, since the image has been smoothed, the covariance matrix between different pixels is not zero. Therefore, there is an uncertianty associated with any of these local maxima. The centroid is then estimated by finding the intersection of two curves, one fitted to three maxima corresponding to rows, and the other fitted to three maxima corresponding to columns.
\section{Simulations and Tests}\label{sec:data}

In order to test the accuracy of the two models and compare them against each other, we need to simulate a large set of stars for which we know the exact position of centroids. Furthermore, we uniformly draw the centroids within sub-pixel regions.

We use the Moffat profile \citep{moffat} for our simulations. Moffat profile is an analytic model for stellar PSFs. It has broader wings than a simple Gaussian model. The surface brightness of the Moffat profile is given by
\beq
I(r) = \frac{F(\beta -1)}{\pi \alpha^{2}}[1+(r/\alpha)^{2}]^{-\beta},
\label{mof}
\eeq
where $F$ is the total flux, $\beta$ is a dimensionless paramter, and $\alpha$ is the length scale of the Moffat profile, with FWHM (hereafter denoted by $\gamma$) being $2\alpha\sqrt{2^{1/\beta}-1}$. 
We want to investigate the performance of the two models introduced in section \ref{sec:model} for different background noise levels, and also different values of $\gamma$. For further simplicity, we set the flux of all stars in our simulations to unity. Per pixel uncertainties are assumed to be uncorrelated Gaussian, and we only study the sky dominated regime.

Moreover, it is more convenient to work with the signal-to-noise ratio (hereafter denoted by SNR) instead of the variance of the Gaussian noise. This is due to the fact that at the same noise level, simulated stars with different values of $\gamma$ have different SNRs.

First, let us find an analytic expression for SNR. We define SNR as the ratio of the mean and variance of the distribution which the flux estimator is drawn from. Assume that the total flux from the point source is $f$, and that the PSF at the $i$-th pixel is given by $P_{i}$. Therefore the brightness of the $i$-th pixel is drawn from a Gaussian distribution $p(I_{i}) = \mathcal{N}(fP_{i},\sigma^{2})$.

The optimal estimator of flux (an estimator which saturates Cramer-Rao bound), is the weighted average $\tilde{f}=\sum_{i}I_{i}P_{i}$. Knowing the distribution of $I_{i}$, it can be shown that 
\beq
p(\tilde{f}) = \mathcal{N}(f , \frac{\sigma^{2}}{\sum_{i}P_{i}^{2}}),
\eeq  
which leads us to
\beq
\begin{array}{l}
\text{SNR} $=$ \frac{\sqrt{\sum_{i} P_{i}^{2}}}{\sigma}.
\end{array}
\label{snr}
\eeq

Note that (\ref{snr}) is valid in the limit of sky-dominated images, and that is the limit we are interested in. In the case of our simulations, the expression (\ref{snr}) can be analytically expressed in terms of $\sigma$, $\gamma$, and also $\beta$. We fix $\beta$ at the fiducial value of 2.5. In this case we have
\beq
\text{SNR} = \frac{0.478}{\sigma \gamma}.
\label{snr2}
\eeq

Equation (\ref{snr2}) implies that for stars with the same flux $f$ and background noise level $\sigma$, those with bigger effective size have lower SNR.
In this work, we study the relation between error in centroid measurement and $\gamma$ at fixed SNR, and vice versa. In the first experiment, we generate 100000 Moffat profiles centered randomly in the sub-pixel region inside the central pixel of a 17$\times$17 image, at four different full width at half maxima of 2, 2.8, 4, and 5.6 pixels with SNR ranging from 5 to 150. In the second experiment, we generate the same number of Moffat profiles with $\gamma$ ranging from 2 pixels to 6 pixels, with four different SNR values of 5, 50, 100, 150.

In addition to the centroiding methods discussed in the previous sections, we compute the error from fitting the PSF model with the exact $\gamma$. That is, we find the best estimates of flux and centroid by optimizing the $\chi^{2}$.

>>>>

If we apply this procedure to every column and row of the 3$\times$3 patch
around the brightest pixel of the smoothed image, we will find three local
maxima for three rows $\{(x_{-},-1)$, $(x_{0},0)$, $(x_{+},1)\}$, and three
local maxima for three columns $\{(-1,y_{-})$, $(0,y_{0})$, $(1,y_{+})\}$.
The centroid is then estimated by finding the intersection of two curves,
one fitted to three maxima corresponding to rows, and the other fitted
to three maxima corresponding to columns.

\section{Simulations and Tests}\label{sec:data}

In order to test the accuracy of the method mentioned so far, and to ompare them against each other,
and we need to simulate a large set of stars for which we know the exact position of centroids.
Furthermore, we uniformly draw the centroids within sub-pixel regions. 
We use the Moffat profile \citep{moffat} for our PSF simulations. 
Moffat profile is an analytic model for stellar PSFs. It has broader wings than
a simple Gaussian model. The surface brightness of the Moffat profile is given by
\beq
I(r) = \frac{F(\beta -1)}{\pi \alpha^{2}}[1+(r/\alpha)^{2}]^{-\beta},
\label{mof}
\eeq
where $F$ is the total flux, $\beta$ is a dimensionless paramter, and $\alpha$ is
the length scale of the Moffat profile, with FWHM (hereafter denoted by $\gamma$)
being $2\alpha\sqrt{2^{1/\beta}-1}$. 
We want to investigate the performance of centroiding methods introduced
in section \ref{sec:model} for different background noise levels, and also different
values of $\gamma$. For further simplicity, we set the flux of all stars in our
simulations to unity. Per pixel uncertainties are assumed to be uncorrelated Gaussian,
and we only study the sky-dominated regime.

Moreover, it is more convenient to work with the signal-to-noise ratio
(hereafter denoted by SNR) instead of the variance of the Gaussian noise.
We define SNR as the ratio of the mean and variance of the distribution
which the flux estimator is drawn from. Assume that the total flux from
the point source is $f$, and that the PSF at the $i$-th pixel is given
by $P_{i}$. Therefore the brightness of the $i$-th pixel is drawn from
a Gaussian distribution $p(I_{i}) = \mathcal{N}(fP_{i},\sigma^{2})$.

The optimal estimator of flux (one which saturates Cramer-Rao bound),
is the weighted average $\tilde{f}=\sum_{i}I_{i}P_{i}$. Knowing the
distribution of $I_{i}$, it can be shown that 
\beq
p(\tilde{f}) = \mathcal{N}(f , \frac{\sigma^{2}}{\sum_{i}P_{i}^{2}}),
\eeq  
which leads us to
\beq
\begin{array}{l}
\text{SNR} $=$ \frac{\sqrt{\sum_{i} P_{i}^{2}}}{\sigma}.
\end{array}
\label{snr}
\eeq

Note that (\ref{snr}) is valid in the limit of sky-dominated regimes,
and that is the limit we are interested in. In the case of our simulations,
SNR given in (\ref{snr}) can be analytically expressed in terms of the per pixel uncertainty
$\sigma$, FWHM $\gamma$, and also $\beta$. We fix $\beta$ at the fiducial value of 2.5.
In this case we have
\beq
\text{SNR} = \frac{0.478}{\sigma \gamma}.
\label{snr2}
\eeq

Equation (\ref{snr2}) implies that for stars with the same flux $f$ and background
noise level $\sigma$, those with radius have lower SNR.

We aim to study how centroiding errors from different methods behave with
changing SNR and size, and how close they get to saturating the CRLB. 
Given the PSF model that we have adopted for our simulations, We need 
to derive an expression for the CRLB as a function of the size of of
the stars, and SNR. We know that the CRLB is given by the inverse of 
the Fisher information matrix \mathcal{F}. The closer an estimator is
to saturating the CRLB, the more information about the quantity that we 
need to estimate is preserved. A measure for closeness of an estimator to 
saturating the CRLB is comparison of the root-mean-squared-error of that estimator 
with the CRLB.

In order to find the CRLB, it is sufficient to find the Fisher matrix.
Let us assume that we have $M$ observables $\{f_{1}, ... , f_{M}\}$, each
related to $B$ model parameters $\theta_{1} , ... , \theta_{B}$. Assuming
un-correlated Gaussian error $\sigma_{m}$ for each observable $f_{m}$, elements
of the Fisher matrix $\mathcal{F}_{ij}$ are given by
\beq
\mathcal{F}_{ij} = \sum_{m}\frac{1}{\sigma_{m}^{2}}\frac{\partial f_{m}}{\partial \theta_{i}}\frac{\partial f_{m}}{\partial \theta_{j}}
\ref{fisher}
\eeq

In our case, the observables are Moffat PSF profile evaluated at different pixels, and  
the model paramters are the centroid coordinates. Thus, \mathcal{F}
is a 2$\times$2 matrix whose elements are given by

\beq
  \mathcal{F} = \sum_{\mathbf{p}}\frac{1}{\sigma_{p}^{2}}
                \frac{\partial f_{\mathbf{p}}}{\partial \theta_{i}}\frac{\partial f_{\mathbf{p}}}{\partial \theta_{j}},
\eeq
where the summation is over pixels, $\theta={x_{c},y_{c}}$, $\sigma_{p}^{2}$ is variance
of the uncorrelated Gaussian noise
\begin{eqnarray}
\langle n(\mathbf{x}) \rangle &=& 0, \\
\langle n(\mathbf{x_{p}})n(\mathbf{x_{p^{\prime}}}) \rangle &=& \sigma^{2}\delta_{\mathbf{p}\mathbf{p}^{\prime}}. 
\end{eqnarray}

In the first experiment, we generate 100000 Moffat profiles centered randomly in the sub-pixel region inside the central pixel of a 17$\times$17 image, at four different full width at half maxima of 2, 2.8, 4, and 5.6 pixels with SNR ranging from 5 to 150. In the second experiment, we generate the same number of Moffat profiles with $\gamma$ ranging from 2 pixels to 6 pixels, with four different SNR values of 5, 50, 100, 150.
In addition to the centroiding methods discussed in the previous sections, we compute the error from fitting the PSF model with the exact $\gamma$. That is, we find the best estimates of flux and centroid by optimizing the $\chi^{2}$.
